### 2014-06-17 kypchak-morphfest-2014/poster.tex
not much left
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@54540 72bbbca6-d526-0410-a7d9-f06f51895060

	\headerbox{Kypchak languages}{name=kypchaklgs,column=0,row=0}{
		\vspace{0.5ex}
		\includegraphics[width=\textwidth]{map/map}
		\vspace{-1.1em}  % -1.25em
		\begin{itemize}
			\item Turkic languages (SOV, agglutinative, vowel harmony)
		\end{itemize}

@@

	\headerbox{Morphological transducers}{name=morphtrans,below=kypchaklgs}{
		\htwo{Morphological transducers}
		\begin{itemize}
			\item Efficient (in speed \& size) models of a language's morphology
			\item Take a surface form, and produce valid lexical form(s)
			\item Take a lexical form, and produce valid surface form(s)\\
		%\end{itemize}
		\scalebox{0.98}{
			алдым \hilitetwo{↔} \texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}, \texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		}
		\end{itemize}

@@

		\htwo{Framework: HFST}
		\begin{itemize}
			\item Reimplements Xerox FST formalisms ({\tt lexc} \& {\tt twol})
			\item Also provides a wrapper around popular free/open-source FST toolkits: SFST, OpenFST, and Foma
		\end{itemize}
		\htwo{Approach}
		\begin{itemize}
			\item morphotactics implemented in \texttt{lexc}
			\item morphophonology implemented in \texttt{twol}
			\item compiled separately; compose-intersected to single transducer\\
			алдым \hilitetwo{↔} \texttt{ал\hilitegrey{>}\archiphon{D}\archiphon{I}\hilitegrey{>}м} \hilitetwo{↔} \texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}
		\end{itemize}

@@

	\headerbox{Development effort}{name=devel,column=4,row=0}{

		\begin{itemize}
			\item Kumyk transducer based on Kazakh, Tatar transducers
			\item \tilda{}1 week to reach 80\% coverage, +1 week to reach 90\%
		\end{itemize}
		\vspace*{-2.5em}
		\hspace*{-2em}\includegraphics[width=0.9\textwidth,angle=270]{../kipchak-lrec-2013/graphs/kumyk-yoldash-cov}
		\vspace*{-2em}
	}

@@

		\headerbox{Categorisation}{name=adjectives,column=4,below=devel}{
		%\headerbox{Morphophonology}{name=morphophon,below=morphtrans}{
			\begin{itemize}
				\item Example:\ morphologically distinct adjective classes (not documented elsewhere)
				\item Other Turkic transducers:\ \texttt{0}-derivation (\hilitetwo{overgenerates})
				\item Our approach:\ categorisation (generates/analyses only correct forms)
			\end{itemize}

		\scalebox{0.72}{
			\begin{tabular}{lllll}
				\toprule
				\textbf{Type} & \textbf{Gloss} & {\small \texttt{\tag{adj}(\tag{comp})}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{subst}}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{advl}}} \\
				\midrule
				\textbf{A1} & \eng{good} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} \\
				\textbf{A2} & \eng{old} & {\qipa иске (искерәк)} & {\qipa иске (искерәк)} & {\qipa — (—)} \\
				\textbf{A3} & \eng{dead} & {\qipa үле (—)} & {\qipa үле (—)} & {\qipa — (—)} \\
				\textbf{A4} & \eng{basic} & {\qipa төп (—)} & {\qipa — (—)} & {\qipa — (—)} \\
				\bottomrule
			\end{tabular}
		}
	}

@@

%			\begin{itemize}
%				\item split into 10 equal parts; coverage calculated over each separately; standard deviation of mean calculated
%			\end{itemize}
			%both split into 10 equal parts; coverage calculated over each separately; standard deviation of the mean was calculated\\
			\htwo{Evaluation measures}
			\begin{itemize}
				\item \textbf{Naïve coverage} - percentage of surface forms in a given corpus receiving ≥ 1 analysis%\\(surface forms may have missing analyses)
				\item \textbf{Mean ambiguity} - average number of analyses for each surface form found in analysed corpus
				\item \textbf{Precision} - of a form's analyses, \% correct
				\item \textbf{Recall} - \% of analyses provided by transducer that are correct for a form, by comparing against a gold standard
			\end{itemize}
			%\htwo{Coverage results}\\
			\htwo{Evaluation results}
%\centering
%\vspace{1pt}\hspace{2em}
			%%\begin{centering}
			%{\large
			%\begin{itemize}
			%	\item each corpus split into 10 equal parts; coverage calculated over each separately; st.\ dev.\ of mean calculated
			%\end{itemize}
			\vspace{-0.5em}
\begin{center}
		\scalebox{0.91}{
			\begin{tabular}{llrrr}
				\toprule
				 & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) & \textbf{Amb.} \\
				\midrule

@@

%			\htwo{Two measures}
%			\htwo{Precision \& recall}
			%\vspace{0.425em}
			\vspace{-0.5em}
			\begin{itemize}
				\item {}{\small selected \& proofed unique random surface forms from news corpora}
			\end{itemize}

			\vspace{-1em}
			\begin{center}
				\begin{tabular}{lrrr}
				\toprule
					\textbf{Language} & \textbf{Forms} & \textbf{Precision} (\%) & \textbf{Recall} (\%) \\
				\midrule
					Kyrgyz & 500 &  &  \\
					Kazakh & 1000 & 98.61 &  57.98 \\
					Tatar & 1000 & 95.03 & 85.65 \\
					Kumyk & 500 & 96.57 & 69.11 \\
				\bottomrule
				\end{tabular}
			\end{center}

@@

			\htwo{Acronyms and numerals}
				\begin{itemize}
					\item \texttt{twol} rules handle phonology for spelt-out words\\
					{\qipa от\hiliteone{ыз}\hilitetwo{да}н} \eng{from thirty}, {\qipa б\hiliteone{ес}\hilitetwo{те}н} \eng{from five}
					\item no phonological triggers available in numerals
					\item (incorrect phonological triggers in acronyms)\\
					{\qipa 30-\hilitetwo{да}н} \eng{from 30}, {\qipa 5-\hilitetwo{те}н} \eng{from 5}
					\item solution:\ phonology-triggering characters
					\item simplified:\ e.g., \archiphon{с} for all voiceless ostruents
				\end{itemize}
				\codeex{
					\begin{spacing}{0.9}
					\texttt{4:4\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "төрт"\\
					5:5\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "бес"\\
					3\%0:3\%0\%\{а\%\}\%\{з\%\}       NUM-DIGIT ; ! "отыз"}\end{spacing}
					\vspace{0em}
				}\\

@@

			\htwo{Related issues}
				\begin{itemize}
					\item Stem-final й combines with following а, о, у to form я, ё, ю
					\item 
				\end{itemize}

@@

		\headerbox{Further information}{name=getting,column=0,below=morphtrans}{
			\begin{itemize}
				\item Part of \textbf{Apertium Turkic} project:\\
				{\small \url{http://wiki.apertium.org/wiki/Apertium\_Turkic}}
				\item Transducers available \textbf{live} at \href{http://turkic.apertium.org/}{\texttt{turkic.apertium.org}}
				\item \textbf{Source code} available from Apertium's svn repo
				%info at {\small \url{http://wiki.apertium.org/wiki/apertium-kir}}

@@

%				\item Trade-off:
				\begin{itemize}
					\item underlying \tag{px3sp} form used:\ \texttt{\archiphon{s}\archiphon{I}\archiphon{n}}
					\item phonological rules delete \texttt{\archiphon{s}}, \texttt{\archiphon{n}} by context
					\item morphophonology more complicated, morphotactics simpler
%				\end{itemize}
			\end{itemize}

>>>>

	\headerbox{Kypchak languages}{name=kypchaklgs,column=0,row=0}{
		\vspace{0.1ex}
		\includegraphics[width=\textwidth]{map/map}
		\vspace{-1.25em}  % -1.25em
		\begin{itemize}
			\item Turkic languages (SOV, agglutinative, vowel harmony)
		\end{itemize}

@@

	\headerbox{Morphological transducers}{name=morphtrans,below=kypchaklgs}{
		\htwo{Morphological transducers}
		\begin{itemize}
			\item Efficient (in speed \& size) models of a language's morphology
			\item Take a surface form, and produce valid lexical form(s)
			\item Take a lexical form, and produce valid surface form(s)\vspace{-0.1ex}\\
		%\end{itemize}
		\scalebox{0.98}{
			алдым \hilitetwo{↔} \texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}, \texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		}
		\end{itemize}

@@

		\htwo{Framework: HFST}
		\begin{itemize}
			\item Reimplements Xerox FST formalisms ({\tt lexc} \& {\tt twol})
			\item Also provides a wrapper around popular free/open-source FST toolkits: SFST, OpenFST, and Foma
		\end{itemize}
		\htwo{Approach}
		\begin{itemize}
			\item morphotactics implemented in \texttt{lexc}
			\item morphophonology implemented in \texttt{twol}
			\item compiled separately; compose-intersected to single transducer\vspace{-0.1ex}\\
			алдым\hfill{}\hilitetwo{↔}\hfill{}\texttt{ал\hilitegrey{>}\archiphon{D}\archiphon{I}\hilitegrey{>}м}\hfill{}\hilitetwo{↔}\hfill{}\texttt{ал\tag{v}\tag{tv}\tag{ifi}\tag{p1}\tag{sg}}\vspace{-0.1ex}\\
			алдым\hfill{}\hilitetwo{↔}\hfill{}\texttt{алд\hilitegrey{>}\archiphon{I}м}\hfill{}\hilitetwo{↔}\hfill{}\texttt{алд\tag{n}\tag{px1sg}\tag{nom}}
		\end{itemize}

@@

	\headerbox{Development effort}{name=devel,column=4,row=0}{
	\scalebox{0.92}{
		\begin{tabular}{lllll}
			\toprule
			 & \textbf{Kyrgyz} & \textbf{Kazakh} & \textbf{Tatar} & \textbf{Kumyk} \\
			\midrule
			begun & Apr.\ 2011 & Dec.\ 2010 & Dec.\ 2011 & Oct.\ 2013 \\
			80\% cov.\ & Aug.?\ 2011 & Aug.\ 2012 & Aug.\ 2012 & Oct.\ 2013 \\\midrule
			time & 4 months & 19 months & 7 months & 1 week \\
			\bottomrule
		\end{tabular}
	}
		\begin{itemize}
			%\item Kyrgyz and Kazakh transducers begun in 2011
			%\item Kyrgyz transducer reached 80\% coverage summer of 2011
			\item Kazakh transducer based on Kyrgyz transducer
			\item Kumyk transducer based on Kazakh, Tatar transducers
			\item \tilda{}1 week to reach 80\% coverage, +1 week to reach 90\%
		\end{itemize}
		\vspace*{-2.5em}
		\hspace*{-2em}\includegraphics[width=0.9\textwidth,angle=270]{../kipchak-lrec-2013/graphs/kumyk-yoldash-cov}
		\vspace*{-2em}
	}

@@

		\headerbox{Categorisation}{name=adjectives,column=4,below=devel}{
		%\headerbox{Morphophonology}{name=morphophon,below=morphtrans}{
			\htwo{Adjectives}
			\begin{itemize}
				\item morphologically distinct adjective classes
				\item most sources:\ adjectives can be used substantively and adverbially
				\item Other Turkic transducers:\ \texttt{0}-derivation (\hilitetwo{overgenerates})
				\item but not all adjectives have all of the following:\\
				comparative forms, substantive readings, adverbial readings
				\item Our approach:\ categorisation 
				\item only correct forms are analysed and generated
			\end{itemize}

		\scalebox{0.7}{
			\begin{tabular}{lllll}
				\toprule
				\textbf{Type} & \textbf{Gloss} & {\small \texttt{\tag{adj}(\tag{comp})}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{subst}}} & {\small \texttt{\tag{adj}(\tag{comp})\tag{advl}}} \\
				\midrule
				\textbf{A1} & \eng{good} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} & {\qipa яхшы (яхшырак)} \\
				\textbf{A2} & \eng{old} & {\qipa иске (искерәк)} & {\qipa иске (искерәк)} & {\qipa — (—)} \\
				\textbf{A3} & \eng{dead} & {\qipa үле (—)} & {\qipa үле (—)} & {\qipa — (—)} \\
				\textbf{A4} & \eng{basic} & {\qipa төп (—)} & {\qipa — (—)} & {\qipa — (—)} \\
				\bottomrule
			\end{tabular}
		}
	}

@@

%			\begin{itemize}
%				\item split into 10 equal parts; coverage calculated over each separately; standard deviation of mean calculated
%			\end{itemize}
			%both split into 10 equal parts; coverage calculated over each separately; standard deviation of the mean was calculated\\
			\htwo{Evaluation measures}
			\begin{itemize}
				\item \textbf{Naïve coverage} - percentage of surface forms in a given corpus receiving ≥ 1 analysis%\\(surface forms may have missing analyses)
				\item \textbf{Mean ambiguity} - average number of analyses for each surface form found in analysed corpus
				\item \textbf{Precision} - of a form's analyses, \% correct
				\item \textbf{Recall} - \% of analyses provided by transducer that are correct for a form, by comparing against a gold standard
			\end{itemize}
			%\htwo{Coverage results}\\
			\htwo{Evaluation results}
%\centering
%\vspace{1pt}\hspace{2em}
			%%\begin{centering}
			%{\large
			%\begin{itemize}
			%	\item each corpus split into 10 equal parts; coverage calculated over each separately; st.\ dev.\ of mean calculated
			%\end{itemize}
			\vspace{-0.5em}
\begin{center}
		\scalebox{0.94}{
			\begin{tabular}{llrrr}
				\toprule
				 & \textbf{Corpus} & \textbf{Tokens} & \textbf{Coverage} (\%) & \textbf{Amb.} \\
				\midrule

@@

%			\htwo{Two measures}
%			\htwo{Precision \& recall}
			%\vspace{0.425em}
			\vspace{-0.4em}
			\begin{itemize}
				\item {}{\small selected \& proofed unique random surface forms from news corpora}
			\end{itemize}

			\vspace{-1em}
			\begin{center}
			\scalebox{1.05}{
				\begin{tabular}{lrrr}
				\toprule
					\textbf{Language} & \textbf{Forms} & \textbf{Precision} (\%) & \textbf{Recall} (\%) \\
				\midrule
					Kyrgyz & 500 &  &  \\
					Kazakh & 1000 & 98.61 &  57.98 \\
					Tatar & 1000 & 95.03 & 85.65 \\
					Kumyk & 500 & 96.57 & 69.11 \\
				\bottomrule
				\end{tabular}
			}
			\end{center}

@@

			\vspace{-0.75ex}
			\htwo{Acronyms and numerals}
				\begin{itemize}
					\item \texttt{twol} rules handle phonology for spelt-out words\\
					{\qipa от\hiliteone{ыз}\hilitetwo{да}н} \eng{from thirty}, {\qipa б\hiliteone{ес}\hilitetwo{те}н} \eng{from five}
					\item no phonological triggers available in numerals\vspace{-0.2em}\\
					 (incorrect phonological triggers in acronyms)\\
					{\qipa 30-\hilitetwo{да}н} \eng{from 30}, {\qipa 5-\hilitetwo{те}н} \eng{from 5}
					\item solution:\ phonology-triggering characters
					\item simplified:\ e.g., \archiphon{с} for all voiceless ostruents
				\end{itemize}
				\codeex{
					\begin{spacing}{0.9}
					\texttt{4:4\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "төрт"\\
					5:5\%\{э\%\}\%\{с\%\}           NUM-DIGIT ; ! "бес"\\
					3\%0:3\%0\%\{а\%\}\%\{з\%\}       NUM-DIGIT ; ! "отыз"}\end{spacing}
					\vspace{0em}
				}\\

			\vspace{-0.75ex}
			\htwo{й + vowel letters}
				\begin{itemize}
					%\item Stem-final й combines with following а, о, у to form я, ё, ю
					\item{} [ а о у ] become [ я ё ю ] after й and й deletes
					\item й incorporated into the context of many rules
					\item additional rules to change the characters and delete original й
				\end{itemize}
%	\vspace{-1ex}
%
				\codeex{
					\begin{spacing}{0.9}
					\texttt{"Deletion of й before yoticised vowels"\\
					й:0 <=> \blank{1em} [ :YotVow ]/[ :0 | \%>:\ ]\ ;}
					\end{spacing}
					\vspace{0em}
				}
%					\item 
%				\end{itemize}

@@

				}
		}

@@

		\headerbox{Further information}{name=getting,column=0,below=morphtrans}{
			\begin{itemize}
				\item Part of \textbf{Apertium Turkic} project:\\
				\url{http://wiki.apertium.org/wiki/Apertium\_Turkic}
				\item Transducers available \textbf{live} at \href{http://turkic.apertium.org/}{\texttt{turkic.apertium.org}}
				\item \textbf{Source code} available from Apertium's svn repo
				%info at {\small \url{http://wiki.apertium.org/wiki/apertium-kir}}

@@

%				\item Trade-off:
				\begin{itemize}
					\item underlying \tag{px3sp} form used:\ \texttt{\archiphon{s}\archiphon{I}\archiphon{n}}
					\item \texttt{\archiphon{s}} and \texttt{\archiphon{n}} default to с and н; rules map to null by context
					\item morphophonology more complicated, morphotactics simpler
%				\end{itemize}
			\end{itemize}
